Online learning methods for financial time series forecasting

### 2017-12-12 slides/defense-slides.tex
mejoras dePeretti

\documentclass{beamer}
%\usepackage[spanish,activeacute]{babel}
\usetheme{metropolis}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{pmat}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{array}
%\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{booktabs}
\usetikzlibrary{shapes,arrows}
%\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}

@@

%\author[Paola Arce]{Paola Arce \\ Departamento de Inform\'atica, UTFSM \\ Thesis advisor: Luis Salinas}
%\date{December 15, 2017}
%\title[Thesis defense]{Online learning methods for financial time series forecasting}
%\subtitle{PhD Thesis Defense}
%
%\titlegraphic{
%    \tikz[overlay,remember picture]
%        \node[at=(current page.south east), anchor=south east] {
%            \includegraphics[height=1.5cm]{img/DIT}
%        };
%}

\author[Paola Arce]{Paola Arce}% \\ Thesis advisor: Luis Salinas}
\date{December 11, 2017}
\title[Thesis defense]{Fast and adaptive cointegration based model for forecasting financial time series}
\subtitle{Seminario Inform\'atica CCTVal}

@@

\begin{frame}
\frametitle{High Frequency data}
High frequency time series are commonly:
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Non-stationary
\item Irregularly spaced over time.
%\item Heterogeneity
%\item Heteroscedastic
\item Potentially cointegrated
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{figure}
\includegraphics[width=0.35\paperwidth]{img/forecast}
\end{figure}
\end{column}
\end{columns}
\vspace{5mm}
This implies that classical statistical models as such can't be used. Therefore, financial time series forecasting is still a challenge.
\end{frame}

\begin{frame}
\frametitle{Stationarity}
\begin{block}{Strong definition}
A {\color{red}strictly stationary time series} $y_t \in \mathbb{R}$, $\forall t \in \mathbb{Z}$, is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
to that of the time shifted set, more precisely:
\begin{equation*}
{\color{blue}
P\{y_{t_1} \leq
c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace }
\end{equation*}
\noindent $\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}$, where $c_1,\dots,c_L$ are constants.
\end{block}
\end{frame}

@@

\begin{frame}
\frametitle{Irregularly spaced over time}
\centerline{
\includegraphics[width=\paperwidth,height=0.7\paperheight]{img/hft-ticks-zoom.png}
}
\end{frame}

%\begin{frame}
%\frametitle{Heteroscedastic}
%Conditional Heteroscedasticity identifies non-constant volatility.
%Volatility gives a measure of uncertainty about future returns. Many factors affect volatility such as: historical prices, trends, monetary policies and news.
%\begin{figure}
%\includegraphics[width=0.8\paperwidth]{img/vix}
%\end{figure}
%\end{frame}

@@

{\Large
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} \phi_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Intercept}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\mathbf{\epsilon}_t $};
        }}_\text{i.i.d}
\end{equation*}
}
\noindent where $p$ is the number of lags and $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$. The columns of $\boldsymbol\beta$ contains the cointegration vectors and the rows of $\boldsymbol\alpha$ correspond with the adjusted vectors.
\end{frame}

@@

@@

\nocite{arce+salinas2012,icpram15,Arce2017}
\bibliographystyle{apalike}
\bibliography{reference}

@@

@@

>>>>

\documentclass{beamer}
%\usepackage[spanish,activeacute]{babel}
\usetheme[sectionpage=none,progressbar=frametitle]{metropolis}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{pmat}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{array}
%\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{booktabs}
\usetikzlibrary{shapes,arrows}
%\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}
\def\Z{{\mathbb Z }}

\makeatletter
%\setlength{\metropolis@titleseparator@linewidth}{2pt}
\setlength{\metropolis@progressonsectionpage@linewidth}{2pt}
\setlength{\metropolis@progressinheadfoot@linewidth}{2pt}
\makeatother

@@

\author[Paola Arce]{Paola Arce \\ Thesis advisor: Luis Salinas}
\date{December 15, 2017}
\title[Thesis defense]{Online learning methods for financial time series forecasting}
\subtitle{PhD Thesis Defense}
%
%\titlegraphic{
%    \tikz[overlay,remember picture]
%        \node[at=(current page.south east), anchor=south east] {
%            \includegraphics[height=1.5cm]{img/DIT}
%        };
%}
%
%\author[Paola Arce]{Paola Arce}% \\ Thesis advisor: Luis Salinas}
%\date{December 11, 2017}
%\title[Thesis defense]{Fast and adaptive cointegration based model for forecasting financial time series}
%\subtitle{Seminario Inform\'atica CCTVal}

@@

\begin{frame}
\frametitle{High Frequency data}
High frequency time series are commonly:
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Non-stationary
\item Irregularly spaced over time.
%\item Heterogeneity
\item Heteroscedastic
\item Potentially cointegrated
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{figure}
\includegraphics[width=0.35\paperwidth]{img/forecast}
\end{figure}
\end{column}
\end{columns}
\vspace{5mm}
This implies that classical statistical models as such can't be used. Therefore, financial time series forecasting is still a challenge.
\end{frame}

\begin{frame}
\frametitle{Stationarity}
\begin{block}{Strong definition}
A {\color{red}strictly stationary time series} $y_t: r.v$ with values in $\mathbb{R}$ defined for $t \in \mathbb{Z}$, is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical to that of the time shifted set, more precisely:
\begin{equation*}
{\color{blue}
P\{y_{t_1} \leq
c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace }
\end{equation*}
\noindent $\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}$, where $c_1,\dots,c_L$ are constants.
\end{block}
\end{frame}

@@

\begin{frame}
\frametitle{Irregularly spaced over time}
\centerline{
\includegraphics[width=\paperwidth,height=0.7\paperheight]{img/hft-ticks-zoom.png}
}
\end{frame}
%
%\begin{frame}
%\frametitle{Heteroscedastic}
%Conditional Heteroscedasticity identifies non-constant volatility.
%Volatility gives a measure of uncertainty about future returns. Many factors affect volatility such as: historical prices, trends, monetary policies and news.
%\begin{figure}
%\includegraphics[width=0.8\paperwidth]{img/vix}
%\end{figure}
%\end{frame}

@@

{\Large
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} {\boldsymbol\phi}_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Intercept}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\boldsymbol{\epsilon}_t $};
        }}_\text{i.i.d}
\end{equation*}
}
\noindent where $\mathbf{y}_t = [y^1_t,\dots,y^l_t]^\intercal$, $p$ is the number of lags and $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$. The columns of $\boldsymbol\beta$ contains the cointegration vectors and the rows of $\boldsymbol\alpha$ correspond with the adjusted vectors.
\end{frame}

@@

\begin{frame}
\frametitle{Articles and conferences}
\nocite{arce+salinas2012,icpram15,Arce2017}
\bibliographystyle{apalike}
\bibliography{reference}
\end{frame}

@@

@@

\begin{frame}
\frametitle{Stationarity}
A time series $y_t: \Z \to \R $ is said to be strictly stationary if, for a finite set of $L$ indices $\{t_{1},\dots,t_{L}\}  \, \forall L \in \Z_{\ge 0}$, the joint probability distribution of the random variables $\{y_{t_{1}},y_{t_{2}},\dots,y_{t_{L}}\}$ remains unchanged for any time shifted $h$.
The probability distribution for the random variable $y_{t_1}$ is:

\begin{equation}
P\{y_{t} \leq c_1 \} = P\{t \in \Z:  y_t \in  ]-\infty, c_1] \} 
\end{equation}

@@

\begin{frame}
\frametitle{Unit root process}
